Estimating Frozen States in Civil Conflicts

### 2024-05-01 paper.qmd
Paper edits

---
title: "Arms Transfers and Frozen Conflict Dynamics"
authors:
- name: J.L.A. Krusell
  email: "joshua.krusell@gu.se"
  affiliations:
      - "University of Gothenburg"
date: today
date-format: long
bibliography: library.bib
linestretch: 1.25
format:
    pdf:
      fontsize: 11pt
      papersize: A4
      include-in-header:
          - text: |
              \usepackage{graphicx}
              \usepackage{tikz}
      pdf-engine: lualatex
execute:
    cache: true
    echo: false
abstract: |
    *Why and when do some frozen conflicts re-escalate back into active conflicts?*
    By definition, frozen conflicts are unresolved conflicts characterized by
    ongoing tensions between belligerents with an ever-present risk of violent
    re-escalation. And yet, there is significant variation in the patterns of
    escalation and de-escalation across frozen conflicts. This paper examines one
    possible factor, international arms transfers to government forces, as a
    possible explainer for why some frozen conflicts are relatively dormant across
    long periods of time, while others undergo periods of intense violent
    instability. To investigate this quantitatively, I use a bespoke, bayesian
    hidden markov model that treats frozen conflicts as a latent, unobserved
    variable and models the transition probability between escalation and
    de-escalation using a set of observed covariates. Preliminary results indicate
    that arms transfers are relatively uninformative for measuring the dynamics of
    frozen conflicts. __Word Count__: 2715
---

@@

@@

library(cmdstanr)
library(dplyr)

log_normalize <- \(x) (x - min(x)) / (max(x) - min(x))

dataset <- readRDS("./data/merge_data.rds") |>
    filter(high_intensity)

@@

Z_prob <- fit$draws("Z_prob", format = "matrix")
active_state <- Z_prob[, sprintf("Z_prob[%d,2]", 1:nrow(dataset))]

>>>>

---
title: "Estimating Frozen States in Civil Conflicts"
subtitle: "A Latent Variable Approach"
authors:
- name: J.L.A. Krusell
  email: "joshua.krusell@gu.se"
  affiliations:
      - "University of Gothenburg"
date: today
date-format: long
bibliography: library.bib
linestretch: 1.25
format:
    pdf:
      fontsize: 11pt
      papersize: A4
      include-in-header:
          - text: |
              \usepackage{graphicx}
              \usepackage{tikz}
      pdf-engine: lualatex
execute:
    cache: true
    echo: false
abstract: |
    *Why and when do some frozen conflicts re-escalate back into active conflicts?*
    By definition, frozen conflicts are unresolved conflicts characterized by
    ongoing tensions between belligerents with an ever-present risk of violent
    re-escalation. And yet, there is significant variation in the patterns of
    escalation and de-escalation across frozen conflicts. This paper examines one
    possible factor, international arms transfers to government forces, as a
    possible explainer for why some frozen conflicts are relatively dormant across
    long periods of time, while others undergo periods of intense violent
    instability. To investigate this quantitatively, I use a bespoke, bayesian
    hidden markov model that treats frozen conflicts as a latent, unobserved
    variable and models the transition probability between escalation and
    de-escalation using a set of observed covariates. Preliminary results indicate
    that arms transfers are relatively uninformative for measuring the dynamics of
    frozen conflicts. __Word Count__: 2715
---

@@

## Predicted vs actual deaths

@@

library(cmdstanr)
library(dplyr)
library(ggplot2)
library(tidyr)

dataset <- readRDS("./data/merge_data.rds")

@@

Zprob <- fit$draws("Gamma", format = "matrix") |> apply(2, median)
K <- fit$metadata()$stan_variable_sizes$beta[1]
for (i in 1:K) {
    state_regex <- \(i) sprintf("^Gamma\\[\\d+,%d\\]$", i)
    dataset[paste0("Zprob", i)] <- exp(Zprob[grepl(state_regex(i), names(Zprob))])
}

dataset$yhat <- fit$draws("yhat", format = "matrix") |> apply(2, median)

df <- select(dataset, unit_id, year, brd, yhat, matches("Zprob")) |>
    pivot_longer(cols = starts_with("Zprob"), names_to = "Z", values_to = "prob") |>
    mutate(Z = sub("^Zprob", "", Z) |> factor(levels = 3:1)) |>
    group_by(unit_id, year) |>
    mutate(prob = prob / sum(prob))

sub.df <- filter(df, unit_id == 63)

ggplot(sub.df, aes(year, brd, col = "Observed")) +
    geom_point() +
    geom_point(aes(year, yhat, col = "Predicted")) +
    geom_bar(aes(year, prob, fill = Z), stat = "identity", position = "stack", alpha = 0.6)

max_observed <- max(sub.df$brd)
scaling_factor <- 1 / max_observed

# Plot the stacked bar plot with the secondary y-axis
ggplot() +
  # Stacked bar plot (background)
  geom_tile(data = sub.df, aes(x = year, y = prob, fill = Z),
           stat = "identity", position = "stack", alpha = 0.2, width = 1) +
  # Observed count as points (overlay, scaled down to match y-axis)
    geom_point(data = sub.df, aes(x = year, y = brd * scaling_factor, col = "Observed")) +
    geom_point(data = sub.df, aes(year, yhat * scaling_factor, col = "Predicted")) +
  # Primary y-axis (for the probability simplex)
  scale_y_continuous(name = "Probability Simplex", sec.axis = sec_axis(~ . / scaling_factor, name = "Observed Count")
                     )  +
    theme_tufte()
